Processing large surveys with package MRP-KIT

### 2021-07-10 case-study/Mapping_CCES_to_ACS.Rmd
checkpointing; convert haven to datafram

# download dataset from IPUMS
# xml and dat file should have same name, be in same directory
acs_2018_subset_ddi <- read_ipums_ddi("usa_00007.xml")
acs_2018_subset <- read_ipums_micro(acs_2018_subset_ddi)
# save as R objects
save(acs_2018_subset, acs_2018_subset_ddi, file= "acs_2018_subset.RData")
```

@@

The `ipumsr` package uses several [tidyverse](https://www.tidyverse.org/) packages
including [`dplyr`](https://dplyr.tidyverse.org/) and [`haven`](https://haven.tidyverse.org/).
The [`read_ipums_micro` function](http://tech.popdata.org/ipumsr/reference/read_ipums_micro.html)
returns a single [`tbl_df`](https://tibble.tidyverse.org/reference/tbl_df-class.html) data frame, i.e., a "tibble".
```{r}
library(labelled)
library(ipumsr)
load("acs_2018_subset.RData")
str(acs_2018_subset, max.level=2, list.len=6)
dim(acs_2018_subset)
names(acs_2018_subset)
```
We selected variables "RACE" and "HISPAN"; the extract contains two columns for each:
"RACE" and "RACED", and "HISPAN" and "HISPAND".  The "D" columns provide an extremely fine-grained
set of categories for race and hispanic origin.  As these go far beyond the categories from
the CCES which are provided, we ignore them.

The columns of this tibble are  [labelled](https://haven.tidyverse.org/reference/labelled.html) vectors.
Labelled vectors can have text labels associated with specific values.
To see how this works, we compare the values and labels for `AGE`.
The 2018 ACS data spans ages 0 through 96,
of these, only ages 0 and 90 have labels.
The function [`ipums_val_labels`](https://tech.popdata.org/ipumsr/reference/ipums_var_info.html)
allows us to inspect these labels.
```{r}
length(table(acs_2018_subset$AGE))
ipums_val_labels(acs_2018_subset$AGE)
```

The function [`as_factor`](https://haven.tidyverse.org/reference/as_factor.html)
allows us to treat labelled vectors as factors.
See the vignette https://tech.popdata.org/ipumsr/reference/ipums_var_info.html for further information.
```{r}
head(table(acs_2018_subset$AGE))
head(table(as_factor(acs_2018_subset$AGE)))
```

The CCES is a survey of _national adults_, i.e., eligible
voters over the age of 18 but the ACS is a survey over the entire
US population, including non-citizens.
We would like to poststratify to a representative survey population,
i.e., the ACS micro-census, but that subset of the general population
who meet the CCES survey criteria, i.e., elibible voters living in
one of the 50 states or Washington DC.
Therefore,  we drop all records for individuals under the age of 18,
non-citizens, or residing outside of the 50 states or DC.

```{r}
table(as_factor(acs_2018_subset$CITIZEN))
# drop underage, non-citizens
citizen = as_factor(acs_2018_subset$CITIZEN) != "Not a citizen"
over_18 = acs_2018_subset$AGE > 17
state_or_DC = acs_2018_subset$STATEFIP < 52
acs_2018_subset = acs_2018_subset[(over_18 & citizen & state_or_DC) , ]  
```

@@

```{r}
library(mrpkit)
names(table(cces_2018_subset$gender))
ipums_val_labels(acs_2018_subset$SEX)
q_sex <- SurveyQuestion$new(name="sex",
      col_names=c("gender", "SEX"),
      values_map=list("Male" = "Male","Female" = "Female"))
q_sex
```

@@

We examine the state names from both the CCES and ACS; the same labels are used in both surveys.
```{r}
# CCES data is a factor, levels are the labels
inputstate <- levels(cces_2018_subset$inputstate)

# ACS dataframe is haven_labelled, treat as factor, only keep states and DC
STATEFIP_levels <- levels(as_factor(acs_2018_subset$STATEFIP))[1:51]
cat("CCES:", inputstate[1:5], sep=", ");
cat("ACS:", STATEFIP_levels[1:5], sep=", ")
table(inputstate == STATEFIP_levels)
```

We need to create a SurveyQuestion object, as above,  but it would be
extremely tedious to do this as a long list of tag, value pairs:
```
values_map=list("Alabama" = "Alabama", ..., "Wyoming" = "Wyoming")
```
The `values_map` is stored as an [R Pairlists](https://github.com/hadley/r-internals/blob/master/pairlists.md).
This is actually a labelled list, not a list of pairs, i.e.,
the expression `list("tag1" = "value1", "tag2" = "value2")`
creates a list of length 2 where each element is a single named item.
```{r}
foo <- list("tag1" = "value1", "tag2" = "value2")
foo
```
To create this list programmatically,
you must create a list of values and a list of labels,
then set the names of the values list to the labels.
```{r}
# create list, then set names (uppercase)
mapping <- as.list(STATEFIP_levels)
names(mapping) <- toupper(inputstate)
mapping[1:3]
# combine the two steps into a one-liner (lowercase)
mapping = as.list(setNames(STATEFIP_levels, tolower(inputstate)))
mapping[1:3]
```

Having established the values map, we create the corresponding SurveyQuestion object.
We use accessor methods `name()`, `col_names()` and `values()` to check our work.
```{r}
state_mapping = as.list(setNames(STATEFIP_levels, inputstate))
q_state <- SurveyQuestion$new(name = "state",
	col_names = c("inputstate", "STATEFIP"),
	values_map= state_mapping)
q_state$name()
q_state$col_names()
head(q_state$values())
```

@@

The CCES and ACS samples record the education level of the respondant
in variables `educ` and `EDUC`, respectively.
Both the CCES and ACS record the highest level of education completed,
but the CCES records the degree obtained whereas the ACS records the year completed.
```{r}
names(table(cces_2018_subset$educ))
ipums_val_labels(acs_2018_subset$EDUC)
```
We choose to align these categories so that the CCES label "2-year"
maps to the ACS labels for 2 and 3 years of college.
Mpping "Some College" to both 1 year and 3 years of college is also possible,
on the rationale that the goal was a 4-year college degree, and that it wasn't obtained.
The choice depends on both an understanding of the survey data and the particular goals
of your analysis.

@@

This is borne out by the CCES, where the question for variable "race" is
"What racial or ethnic group best describes you?",
and includes "Hispanic" as a possible response.
The ACS variable "RACE" has a different set of responses:
```{r}
names(table(cces_2018_subset$race))
ipums_val_labels(acs_2018_subset$RACE)
```

In order to use the ACS for poststratification, we need to create a variable
corresponding to the CCES variable "race" from a combination of ACS variables.
Both surveys contain variable "Hispanic Origin".
For the CCES, this is a "Yes/No" question:
"Are you of Spanish, Latino, or Hispanic origin or descent?".
In the ACS, the variable HISPAN identifies persons of Hispanic/Spanish/Latino origin and
classifies them according to their country of origin.
```{r}
names(table(cces_2018_subset$hispan))
ipums_val_labels(acs_2018_subset$HISPAN)
```

We propose to add a column to the ACS survey which combines information present in the `RACE` and `HISPAN` variables.
```{r}
# create column which blends "race" "hispan"
acs_race2 <- as_factor(acs_2018_subset$RACE)
# population percentages by label
prop.table(table(acs_race2))
# if "hispan" is "Yes", "race" label is "Hispanic".
is_hispan <- acs_2018_subset$HISPAN > 0 & acs_2018_subset$HISPAN < 5
length(is_hispan[is_hispan == TRUE]) / length(is_hispan)
levels(acs_race2) = c(levels(acs_race2), "Hispanic")
acs_race2[is_hispan==TRUE] <- "Hispanic"
# adjusted percentages
prop.table(table(acs_race2))
# add to acs_2018_subset data.frame - respect column types
acs_2018_subset$race2 = to_labelled(acs_race2)
```

@@

```{r}
names(table(cces_2018_subset$marstat))
ipums_val_labels(acs_2018_subset$MARST)
```

@@

```{r}
# create column which conflates spouse present, absent
acs_marst2 <- as_factor(acs_2018_subset$MARST)
levels(acs_marst2) = c(levels(acs_marst2), "Married")
acs_marst2[acs_marst2 =="Married, spouse present"] <- "Married"
acs_marst2[acs_marst2 =="Married, spouse absent"] <- "Married"

# add to acs_2018_subset data.frame - respect column types
acs_2018_subset$marst2 = to_labelled(droplevels(acs_marst2))
ipums_val_labels(acs_2018_subset$marst2)
```
Now we can create a mapping which maps both "Married" and ""Domestic / civil partnership"
to the the (new) ACS label "Married".^[
Alternatively, we could try to access more ACS variables which account for
same sex marriages and civil unions and combine information across multiple columns, as we did
for ACS variable race2.]

@@

# transform ACS age to birthyr, then add to ACS
acs_2018_subset$birthyr = 2018 - acs_2018_subset$AGE
acs_2018_subset$birthyr_cohort = to_labelled(cut(acs_2018_subset$birthyr, cohort_breaks, cohort_labels))
```

@@

A SurveyData object requires a dataframe, a question list, and a response list.
A question list is a list of tag-value pairs where the tag (name) is the dataframe column name and the value is a description of corresponding survey question.
For the CCES, the dataframe column descriptions are lists in the attrbute "var.labels".
As before, we create this list programatically using the `as.list(setNames(tags,values))` trick.
```{r}
cces_2018_questions <- as.list(setNames(attr(cces_2018_subset,"var.labels"), names(cces_2018_subset)))
head(cces_2018_questions)
```
For the ACS, we use the function `ipumsr::ipums_var_label` to get a short description of the variable.
```{r}
acs_2018_var_labels <- c()
for (name in names(acs_2018_subset)) {
    acs_2018_var_labels <- c(acs_2018_var_labels, ipums_var_label(acs_2018_subset[[name]]))
}
acs_2018_questions <- as.list(setNames(acs_2018_var_labels, names(acs_2018_subset)))
head(acs_2018_questions)
```

A response list is a list of name-value pairs which map the column name to the set of
possible response values.
We can create this list using a for loop, or by using R's `apply` function.
```{r}
# create response list using a for loop
cces_2018_responses <- list()
for (name in names(cces_2018_subset)) {
    cces_2018_responses[[name]] <- levels(as.factor(cces_2018_subset[[name]]))
}
head(cces_2018_responses)

@@

We use these lists to create SurveyData objects for both the CCES and ACS.
```{r}
cces_2018_data <- SurveyData$new(cces_2018_subset, cces_2018_questions, cces_2018_responses)
cces_2018_data
acs_2018_data <- SurveyData$new(acs_2018_subset, acs_2018_questions, acs_2018_responses)
acs_2018_data
```

>>>>

# download dataset from IPUMS
# xml and dat file should have same name, be in same directory
acs_2018_subset_ddi <- read_ipums_ddi("usa_00007.xml")
acs_2018_subset_tbl <- read_ipums_micro(acs_2018_subset_ddi)
# save as R objects
save(acs_2018_subset_tbl, acs_2018_subset_ddi, file= "acs_2018_subset.RData")
```

@@

The `ipumsr` package uses several [tidyverse](https://www.tidyverse.org/) packages
including [`dplyr`](https://dplyr.tidyverse.org/) and [`haven`](https://haven.tidyverse.org/).
The [`read_ipums_micro` function](http://tech.popdata.org/ipumsr/reference/read_ipums_micro.html)
returns a single [`tbl_df`](https://tibble.tidyverse.org/reference/tbl_df-class.html) data frame, i.e., a "tibble".
```{r}
library(labelled)
library(ipumsr)
load("acs_2018_subset.RData")
str(acs_2018_subset_tbl, max.level=2, list.len=6)
dim(acs_2018_subset_tbl)
names(acs_2018_subset_tbl)
```
We selected variables "RACE" and "HISPAN"; the extract contains two columns for each:
"RACE" and "RACED", and "HISPAN" and "HISPAND".  The "D" columns provide an extremely fine-grained
set of categories for race and hispanic origin.  As these go far beyond the categories from
the CCES which are provided, we ignore them.

The columns of this tibble are  [labelled](https://haven.tidyverse.org/reference/labelled.html) vectors.
Labelled vectors can have text labels associated with specific values.
To see how this works, we compare the values and labels for `AGE`.
The 2018 ACS data spans ages 0 through 96,
of these, only ages 0 and 90 have labels.
The function [`ipums_val_labels`](https://tech.popdata.org/ipumsr/reference/ipums_var_info.html)
allows us to inspect these labels.
```{r}
length(table(acs_2018_subset_tbl$AGE))
ipums_val_labels(acs_2018_subset_tbl$AGE)
```

The function [`as_factor`](https://haven.tidyverse.org/reference/as_factor.html)
allows us to treat labelled vectors as factors.
The function [`zap_labels`](https://haven.tidyverse.org/reference/zap_labels.html)
allows us to treat labelled vectors as vectors.
See https://haven.tidyverse.org/articles/semantics.html for further information.
```{r}
head(table(acs_2018_subset_tbl$AGE))
head(table(as_factor(acs_2018_subset_tbl$AGE)))
head(table(zap_labels(acs_2018_subset_tbl$AGE)))
```

The CCES is a survey of _national adults_, i.e., eligible
voters over the age of 18 but the ACS is a survey over the entire
US population, including non-citizens.
We would like to poststratify to a representative survey population,
i.e., the ACS micro-census, but that subset of the general population
who meet the CCES survey criteria, i.e., elibible voters living in
one of the 50 states or Washington DC.  (Note: STATEFIP codes have gaps, Wyoming has STATEFIP code 56).
Therefore,  we drop all records for individuals under the age of 18 and non-citizens, or citizens not
living in the 50 states or DC.
```{r}
# using the level name here
table(as_factor(acs_2018_subset_tbl$CITIZEN))
citizen = as_factor(acs_2018_subset_tbl$CITIZEN) != "Not a citizen"
# using the column value
over_18 = acs_2018_subset_tbl$AGE > 17
state_or_DC = acs_2018_subset_tbl$STATEFIP < 57  
acs_2018_subset_tbl = acs_2018_subset_tbl[(over_18 & citizen & state_or_DC) , ]  
```

The metadata is in the object `acs_2018_subset_ddi`.
Working with the `tbl_df` object is awkward, as the MRP-KIT API expects to work with simple dataframes.
Therefore we create a data.frame object which contains only the columns of interest as factors,
excepting "AGE" which is count, not categorical, data.
```{r}
acs_2018_subset <- data.frame(
    "STATEFIP" = droplevels(as_factor(acs_2018_subset_tbl$STATEFIP)),
    "SEX" = droplevels(as_factor(acs_2018_subset_tbl$SEX)),
    "AGE" = zap_labels(acs_2018_subset_tbl$AGE),
    "MARST" = droplevels(as_factor(acs_2018_subset_tbl$MARST)),
    "RACE" = droplevels(as_factor(acs_2018_subset_tbl$RACE)),
    "HISPAN" = droplevels(as_factor(acs_2018_subset_tbl$HISPAN)),
    "CITIZEN" = droplevels(as_factor(acs_2018_subset_tbl$CITIZEN)),
    "EDUC" = droplevels(as_factor(acs_2018_subset_tbl$EDUC))
)
str(acs_2018_subset, max.level=2, list.len=12, give.attr=FALSE)
```

@@

```{r}
library(mrpkit)
levels(cces_2018_subset$gender)
levels(acs_2018_subset$SEX)
q_sex <- SurveyQuestion$new(name="sex",
      col_names=c("gender", "SEX"),
      values_map=list("Male" = "Male","Female" = "Female"))
q_sex
```

@@

We examine the state names from both the CCES and ACS; the same labels are used in both surveys.
```{r}
cat("CCES:", levels(cces_2018_subset$inputstate)[1:5], "\n", sep=", ");
cat("ACS:", levels(acs_2018_subset$STATEFIP)[1:5], "\n", sep=", ");
table(levels(cces_2018_subset$inputstate) == levels(acs_2018_subset$STATEFIP))
```

We need to create a SurveyQuestion object, as above,  but it would be
extremely tedious to do this as a long list of tag, value pairs:
```
values_map=list("Alabama" = "Alabama", ..., "Wyoming" = "Wyoming")
```
The `values_map` is stored as an [R Pairlists](https://github.com/hadley/r-internals/blob/master/pairlists.md).
This is actually a labelled list, not a list of pairs, i.e.,
the expression `list("tag1" = "value1", "tag2" = "value2")`
creates a list of length 2 where each element is a single named item.
```{r}
foo <- list("tag1" = "value1", "tag2" = "value2")
foo
```
To create this list programmatically,
you must create a list of values and a list of labels,
then set the names of the values list to the labels.
```{r}
state_names <- levels(cces_2018_subset$inputstate)
state_values <- levels(acs_2018_subset$STATEFIP)
# create named list in two steps
mapping <- as.list(state_values)
names(mapping) <- state_names 
mapping[1:3]
# one-liner using `setNames`
mapping = as.list(setNames(state_values, state_names))
mapping[1:3]
```

Having established the values map, we create the corresponding SurveyQuestion object.
We use accessor methods `name()`, `col_names()` and `values()` to check our work.
```{r}
state_mapping = as.list(setNames(state_values, state_names))
q_state <- SurveyQuestion$new(name = "state",
	col_names = c("inputstate", "STATEFIP"),
	values_map= state_mapping)
q_state$name()
q_state$col_names()
head(q_state$values())
```

@@

The CCES and ACS samples record the education level of the respondant
in variables `educ` and `EDUC`, respectively.
Both the CCES and ACS record the highest level of education completed,
but the CCES records the degree obtained whereas the ACS records the year completed.
```{r}
levels(cces_2018_subset$educ)
levels(acs_2018_subset$EDUC)
```
We choose to align these categories so that the CCES label "2-year"
maps to the ACS labels for 2 and 3 years of college.
Mpping "Some College" to both 1 year and 3 years of college is also possible,
on the rationale that the goal was a 4-year college degree, and that it wasn't obtained.
The choice depends on both an understanding of the survey data and the particular goals
of your analysis.

@@

This is borne out by the CCES, where the question for variable "race" is
"What racial or ethnic group best describes you?",
and includes "Hispanic" as a possible response.
The ACS variable "RACE" has a different set of responses:
```{r}
levels(cces_2018_subset$race)
levels(acs_2018_subset$RACE)
```

In order to use the ACS for poststratification, we need to create a variable
corresponding to the CCES variable "race" from a combination of ACS variables.
Both surveys contain variable "Hispanic Origin".
For the CCES, this is a "Yes/No" question:
"Are you of Spanish, Latino, or Hispanic origin or descent?".
In the ACS, the variable HISPAN identifies persons of Hispanic/Spanish/Latino origin and
classifies them according to their country of origin.
```{r}
levels(cces_2018_subset$hispan)
levels(acs_2018_subset$HISPAN)
```

We propose to add a column to the ACS survey which combines information present in the `RACE` and `HISPAN` variables.
```{r}
# create column which blends "race" "hispan"
acs_race2 <- acs_2018_subset$RACE
# population percentages by label
prop.table(table(acs_race2))
# if "hispan" is "Yes", "race" label is "Hispanic".
is_hispan <- acs_2018_subset$HISPAN != "Not Hispanic"
length(is_hispan[is_hispan == TRUE]) / length(is_hispan)
levels(acs_race2) = c(levels(acs_race2), "Hispanic")
acs_race2[is_hispan==TRUE] <- "Hispanic"
# adjusted percentages
prop.table(table(acs_race2))
# add to acs_2018_subset data.frame - respect column types
acs_2018_subset$race2 = acs_race2
```

@@

```{r}
levels(cces_2018_subset$marstat)
levels(acs_2018_subset$MARST)
```

@@

```{r}
# create column which conflates spouse present, absent
acs_marst2 <- acs_2018_subset$MARST
levels(acs_marst2) = c(levels(acs_marst2), "Married")
acs_marst2[acs_marst2 =="Married, spouse present"] <- "Married"
acs_marst2[acs_marst2 =="Married, spouse absent"] <- "Married"

# add to acs_2018_subset data.frame
acs_2018_subset$marst2 = acs_marst2
```
Now we can create a mapping which maps both "Married" and ""Domestic / civil partnership"
to the the (new) ACS label "Married".^[
Alternatively, we could try to access more ACS variables which account for
same sex marriages and civil unions and combine information across multiple columns, as we did
for ACS variable race2.]

@@

# transform ACS age to birthyr, then add to ACS
acs_2018_subset$birthyr = 2018 - acs_2018_subset$AGE
acs_2018_subset$birthyr_cohort = cut(acs_2018_subset$birthyr, cohort_breaks, cohort_labels)
```

@@

A SurveyData object requires a dataframe, a question list, and a response list.
A question list is a list of tag-value pairs where the tag (name) is the dataframe column name and the value is a description of corresponding survey question.
For the CCES, the dataframe column descriptions are lists in the attrbute "var.labels".
As before, we create this list programatically using the `as.list(setNames(tags,values))` trick.
```{r, EVAL=FALSE}
cces_2018_questions <- as.list(setNames(attr(cces_2018_subset,"var.labels"), names(cces_2018_subset)))
head(cces_2018_questions)
```
For the ACS, we use function `ipumsr::ipums_var_label` to get a short description of the variable
from the metadata in the DDI object.
```{r, EVAL=FALSE}
acs_2018_var_labels <- c()
for (name in names(acs_2018_subset)) {
    acs_2018_var_labels <- c(acs_2018_var_labels, ipums_var_label(acs_2018_subset[[name]]))
}
acs_2018_questions <- as.list(setNames(acs_2018_var_labels, names(acs_2018_subset)))
head(acs_2018_questions)
```

A response list is a list of name-value pairs which map the column name to the set of
possible response values.
We can create this list using a for loop, or by using R's `apply` function.
```{r, EVAL=FALSE}
# create response list using a for loop
cces_2018_responses <- list()
for (name in names(cces_2018_subset)) {
    cces_2018_responses[[name]] <- levels(as.factor(cces_2018_subset[[name]]))
}
head(cces_2018_responses)

@@

We use these lists to create SurveyData objects for both the CCES and ACS.
```{r, EVAL=FALSE}
cces_2018_data <- SurveyData$new(cces_2018_subset, cces_2018_questions, cces_2018_responses)
cces_2018_data
acs_2018_data <- SurveyData$new(acs_2018_subset, acs_2018_questions, acs_2018_responses)
acs_2018_data
```
Now we can create the mapping object from the CCES to the ACS.
```{r, EVAL=FALSE}
cces_acs_mapping <- SurveyMap$new(cces_2018_data, acs_2018_data, q_sex, q_state, q_race, q_educ)
# poststrat table size: 2 * 51 * 8 * 6
```
